Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-07 cordillera-cycle.tex
Add last flow direction map in manuscript.
Polish model setup.

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments
\citep{martin-etal-2011,aschwanden-etal-2013}. The till effective pressure is
related to the modelled amount of water in the till, using a formula derived
from laboratory experiments with till extracted from an Antarctic ice stream.
Basal topography is derived from the ETOPO1 combined topography
and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.

Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}.

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr}, excluding variability associated with the
seasonal cycle itself \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
$\{T_m0, P_m0\}$, modified by temperature-offset time series ${\Delta}T_{TS}$
and a lapse-rate correction ${\Delta}T_{LR}$.
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t) + {\Delta}T_{LR}(t, x, y) \\
    P_m(t, x, y) &= P_{m0}(x, y)
\end{align}

The present-day climatology $\{T_m0, P_m0\}$ was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study.

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \aref:
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2]
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided by their authors \aref. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

This correction accounts for the evolution of ice thickness $h$ on the one
hand, and for differences between the climate forcing reference topography and
the ice flow model basal topography $z_{b}-z_{ref}$ on the other hand. An
annual lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$ is used in all
simulations.

@@

>>>>

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments
\citep{martin-etal-2011,aschwanden-etal-2013}. The till effective pressure is
related to the modelled amount of water in the till, using a formula derived
from laboratory experiments with till extracted from an Antarctic ice stream
\aref. Basal topography is derived from the ETOPO1 combined topography
and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}.

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis (NARR) \citep{data:narr}, after excluding variability associated
with the
seasonal cycle itself \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
$\{T_{m0}, P_{m0}\}$, modified by temperature-offset time series
${\Delta}T_{TS}$ and a lapse-rate correction ${\Delta}T_{LR}$.
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t) + {\Delta}T_{LR}(t, x, y) \\
    P_m(t, x, y) &= P_{m0}(x, y)
\end{align}

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from the
near-surface air temperature and precipitation rate fields of the NARR over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study \aref.

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore \aref. Palaeo-temperatures from the GRIP and
NGRIP records were calculated using a quadratic equation \aref:
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2]
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided by their authors \aref. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

This correction accounts for the evolution of ice thickness $h$ on the one
hand, and for differences between the the ice flow model basal topography
$z_{b}$ and the climate forcing reference topography $z_{ref}$ on the other
hand. All simulations use an annual temperature lapse rate of
$\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.

@@

% fig:lastflow
\begin{figure}
  \includegraphics{lastflow}
  \caption{Modelled directions of the last basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout the simulation.
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.}
  \label{fig:lastflow}
\end{figure}
